### 2024-07-09 paper/tex/03-declarative-system.tex
edit and prepare for jfp

\begin{figure}[h]

@@

\subparagraph*{Variables}
    Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
    the type of a variable from the context. 
    In literature, one can find another version of this rule,
    that enables inferring a type \emph{equivalent}
    to the type from the context. 
    In our case, the inference of equivalent types is admissible by \ruleref{\ottdruleDTPEquivLabel}.

\subparagraph*{Annotations}
  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
  The annotation is only valid if the
  inferred type is a subtype of the annotation type.

\subparagraph*{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type $[[iP]]$ annotating the argument is well-formed,
  and then infers the type $[[iN]]$ of the body in the extended context.
  As a result, it returns an arrow type $[[iP → iN]]$.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type $[[iN]]$ of the body. As a result, it returns a polymorphic type
  $[[∀pa.iN]]$.

\subparagraph*{Return and thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  simply add the corresponding shift constructors to the type of the body.

\subparagraph*{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and ensures that the bound variables do not escape the scope.

\subparagraph*{Let binders}
  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
  binders: we infer the type of the bound value and continue the typing of the
  computation in the extended context.

@@

\subparagraph*{Applicative let binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

>>>>

\begin{figure}[t]

@@

\paragraph*{Variables}
    Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
    the type of a variable from the context. 
    In literature, one can find another version of this rule,
    that enables inferring a type \emph{equivalent}
    to the type from the context. 
    In our case, the inference of equivalent types is admissible by \ruleref{\ottdruleDTPEquivLabel}.

\paragraph*{Annotations}
  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
  The annotation is only valid if the
  inferred type is a subtype of the annotation type.

\paragraph*{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type $[[iP]]$ annotating the argument is well-formed,
  and then infers the type $[[iN]]$ of the body in the extended context.
  As a result, it returns an arrow type $[[iP → iN]]$.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type $[[iN]]$ of the body. As a result, it returns a polymorphic type
  $[[∀pa.iN]]$.

\paragraph*{Return and thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  simply add the corresponding shift constructors to the type of the body.

\paragraph*{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and ensures that the bound variables do not escape the scope.

\paragraph*{Let binders}
  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
  binders: we infer the type of the bound value and continue the typing of the
  computation in the extended context.

@@

\paragraph*{Applicative let binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

### 2024-07-09 paper/tex/04-algorithm.tex
edit and prepare for jfp

  For convenience and scalability, 
  we decompose the subtyping algorithm 
  into several procedures. \Cref{fig:alg-subtyping-graph}
  shows these procedures and the dependencies between them:
  arrows denote the invocation of one procedure from another.

@@

\begin{figure}[t]
  \hfill
  \begin{multicols}{2}
  \ottdefnUNUnifLabeled{}
  \columnbreak\\
  \ottdefnUPUnifLabeled{}
  \end{multicols}
  \caption{Unification Algorithm}
  \label{fig:unification}
\end{figure}

@@

Both type upgrade and the least upper bound algorithms are used
to find a minimal supertype under certain conditions. 
For a given type $[[iP]]$ well-formed in $[[Γ]]$, the \emph{upgrade} operation 
finds the least among those supertypes of $[[iP]]$ that are well-formed
in a smaller context $[[Δ ⊆ Γ]]$.
For given two types $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$,
the \emph{least upper bound} operation finds the least among
common supertypes of $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$.
These algorithms are shown in \cref{fig:type-upgrade}.

@@

\paragraph*{The Type Upgarde}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ nor $[[pnbs]]$.

This way, to find the least supertype of $[[iP]]$ well-formed in $[[Δ]] = [[Γ \ {pnas}]]$
(\ie not containing $[[pnas]]$), we can do the following.
First, construct a new type $[[iP']]$ by renaming $[[pnas]]$ in $[[iP]]$ to fresh $[[pnbs]]$,
and second, find \emph{the least upper bound} of $[[iP]]$ and $[[iP']]$ in the appropriate
context. However, for reasons of symmetry, in rule
\ruleref{\ottdruleLUBUpgradeLabel} we employ a different but equivalent approach:
we create \emph{two} types $[[iP1]]$ and $[[iP2]]$ constructed by renaming $[[pnas]]$ in $[[iP]]$
to fresh disjoint variables $[[pnbs]]$ and $[[pncs]]$ respectively, and then 
find the least upper bound of $[[iP1]]$ and $[[iP2]]$.

@@

  In our representation of the anti-unification algorithm, this `merge' happens
  automatically:
  following the rule \ruleref{\ottdruleAUAULabel},
  the name of the `hole' is uniquely defined by the pair of types it is mapped to.  
  Specifically, when anti-unifying $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$ our algorithm returns 
  $[[↓α̂⁻_{α⁻, β⁻} → α̂⁻_{α⁻, β⁻}]]$, that is a renaming of $[[↓γ̂⁻ → γ̂⁻]]$.

  This way, as the output the rule returns the following tuple:
  \begin{itemize}
    \item $[[Ξ1 ∪ Ξ2]]$---a simple union of the sets of `holes' 
      returned from by the recursive calls, 
    \item $[[uQ → uM]]$---the resulting pattern 
      constructed from the patterns returned recursively.
    \item $[[aus1 ∪ aus1']]$ and $[[aus2 ∪ aus'2]]$
      --- a union (in a relational sense)
      of the substitutions returned by the recursive calls. 
      It is worth noting that the union is well-defined because
      the result of the substation on a `hole' is determined by the 
      name of the `hole'.
  \end{itemize}

@@

  Third, the rule is only applicable if all other rules fail. Notice that it
  could happen even when the input types have matching constructors. For
  example, the generalizer of $[[↑α⁺]]$ and $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with
  mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$), rather than $[[↑γ̂⁺]]$. This
  way, the algorithm must try to apply the congruent rules first, and only if
  they fail, apply \ruleref{\ottdruleAUAULabel}. As it is not known a priori
  whether the congruent rules will be applicable, the implementation must use 
  backtracking.

@@

The positive and the negative type inference judgments have symmetric forms:
$[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ c : iN]]$. Both of these algorithms
take as an input typing context $[[Γ]]$, a variable context $[[Φ]]$, and 
a term (a value or a computation) taking its type variables from $[[Γ]]$, 
and term variables from $[[Φ]]$. As an output, they return a type
of the given term, which we guarantee to be normalized.

The application type inference judgment has form 
$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$.
As an input, it takes three contexts: typing context $[[Γ]]$, a variable context $[[Φ]]$,
and an instantiation context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
a list of arguments (terms) $[[args]]$ the head is applied to.
The head may contain algorithmic variables specified by $[[Θ1]]$, 
in other words, $[[Γ; dom(Θ1) ⊢ uN]]$.
As a result, the application inference judgment returns 
$[[uM]]$---a normalized type of the result of the application.
Type $[[uM]]$ may contain new algorithmic variables, and thus, 
the judgment also returns $[[Θ2]]$---an updated instantiation context
and $[[SC]]$---a set of subtyping constraints.
Together $[[Θ2]]$  and $[[SC]]$ specify how the algorithmic variables 
must be instantiated.

\begin{figure}[t]
  \ottdefnATPInfLabeled{}
  \hfill \\
  \ottdefnATNInfLabeled{}
  \hfill \\
  \ottdefnATSpinInfLabeled{}
  \caption{Algorithmic Type Inference}
  \label{fig:type-inference}
\end{figure}

@@

\paragraph*{Application to an Empty List of Arguments}
  Rule \ruleref{\ottdruleATEmptyAppLabel}
  is the base case of application inference. 
  If the list of applied arguments is empty, 
  the inferred type is the type of the head,
  and the algorithm returns it after normalizing.

@@

To find this substitution, we use the minimal instantiation
algorithm (\cref{fig:minimal-instantiation}). The judgment `$[[Γ ⊢ uP SC minby uσ]]$'
$[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
instantiation of $[[uP]]$ that satisfies $[[SC]]$. 
First, it removes the existential quantifiers by \ruleref{\ottdruleSINGExistsLabel}.
Then it considers two cases. 
\begin{enumerate}
  \item If the type $[[uP]]$ is an algorithmic variable $[[α̂⁺]]$
    restricted by a \emph{subtyping} constraint $[[(α̂⁺ :≥ iQ) ∊ SC]]$,
    its minimal instantiation is the type $[[nf(iQ)]]$, and 
    \ruleref{\ottdruleSINGPUvarLabel} is applied.
  \item Otherwise, the type $[[uP]]$ is either a shifted computation, a
    declarative variable, an unrestricted algorithmic variable, or an
    algorithmic variable restricted by an \emph{equivalence} constraint. In all
    of these cases, the minimal instantiation exists if and only if there is
    only one possible instantiation of $[[uP]]$ that satisfies $[[SC]]$.

>>>>

  We decompose the subtyping algorithm 
  into several procedures. \Cref{fig:alg-subtyping-graph}
  shows these procedures and the dependencies between them:
  arrows denote the invocation of one procedure from another.

@@

\begin{figure}[t]
  \begin{minipage}[t]{0.50\textwidth}
    \ottdefnUNUnifLabeled{}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.49\textwidth}
    \ottdefnUPUnifLabeled{}
  \end{minipage}
  \caption{Unification Algorithm}
  \label{fig:unification}
\end{figure}

@@

Both type upgrade and the least upper bound algorithms are used
to find a minimal supertype under certain conditions. 
For a given type $[[iP]]$ well-formed in $[[Γ]]$, the \emph{upgrade} operation 
finds the least among those supertypes of $[[iP]]$ that are well-formed
in a smaller context $[[Δ ⊆ Γ]]$.
When given two types $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$,
the \emph{least upper bound} operation finds the least among
common supertypes of $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$.
These algorithms are shown in \cref{fig:type-upgrade}.

@@

\paragraph*{The Type Upgrade}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ or $[[pnbs]]$.

This way, to find the least supertype of $[[iP]]$ well-formed in $[[Δ]] = [[Γ \ {pnas}]]$
(\ie not containing $[[pnas]]$), we can do the following.
First, construct a new type $[[iP']]$ by renaming $[[pnas]]$ in $[[iP]]$ to fresh $[[pnbs]]$,
and second, find \emph{the least upper bound} of $[[iP]]$ and $[[iP']]$ in the appropriate
context. However, for reasons of symmetry, in rule
\ruleref{\ottdruleLUBUpgradeLabel} we employ a different but equivalent approach:
we create \emph{two} types $[[iP1]]$ and $[[iP2]]$ by renaming $[[pnas]]$ in $[[iP]]$
to fresh disjoint variables $[[pnbs]]$ and $[[pncs]]$ respectively, and then 
find the least upper bound of $[[iP1]]$ and $[[iP2]]$.

@@

  In our representation of the anti-unification algorithm, this `merging' happens
  automatically:
  following the rule \ruleref{\ottdruleAUAULabel},
  the name of the `hole' is uniquely defined by the pair of types it is mapped to.  
  Specifically, when anti-unifying $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$ our algorithm returns 
  $[[↓α̂⁻_{α⁻, β⁻} → α̂⁻_{α⁻, β⁻}]]$, that is a renaming of $[[↓γ̂⁻ → γ̂⁻]]$.

  This way, as the output the rule returns the following tuple:
  \begin{itemize}
    \item $[[Ξ1 ∪ Ξ2]]$---a simple union of the sets of `holes' 
      returned from by the recursive calls, 
    \item $[[uQ → uM]]$---the resulting pattern 
      constructed from the patterns returned recursively.
    \item $[[aus1 ∪ aus1']]$ and $[[aus2 ∪ aus'2]]$
      --- a union (in a relational sense)
      of the substitutions returned by the recursive calls. 
      It is worth noting that the union is well-defined because
      the result of the substitution on a `hole' is determined by the 
      name of the `hole'.
  \end{itemize}

@@

  Third, the rule is only applicable if all other rules fail. Notice that it
  could happen even when the input types have matching constructors. For
  example, the generalizer of $[[↑α⁺]]$ and $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with
  mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$), rather than $[[↑γ̂⁺]]$. This
  way, the algorithm must try to apply the congruent rules first, and only if
  they fail, apply \ruleref{\ottdruleAUAULabel}.

@@

The positive and the negative type inference judgments have the symmetric forms:
$[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ c : iN]]$. Both of these algorithms
take as an input typing context $[[Γ]]$, a variable context $[[Φ]]$, and 
a term (a value or a computation) taking its type variables from $[[Γ]]$, 
and term variables from $[[Φ]]$. As an output, they return a type
of the given term, which we guarantee to be normalized.

The application type inference judgment has the form 
$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$.
As an input, it takes three contexts: typing context $[[Γ]]$, a variable context $[[Φ]]$,
and an instantiation context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
a list of arguments (terms) $[[args]]$ the head is applied to.
The head may contain algorithmic variables specified by $[[Θ1]]$, 
in other words, $[[Γ; dom(Θ1) ⊢ uN]]$.
As a result, the application inference judgment returns 
$[[uM]]$---a normalized type of the result of the application.
Type $[[uM]]$ may contain new algorithmic variables, and thus, 
the judgment also returns $[[Θ2]]$---an updated instantiation context
and $[[SC]]$---a set of subtyping constraints.
Together $[[Θ2]]$ and $[[SC]]$ specify how the algorithmic variables 
must be instantiated.

\begin{figure}[t!]
  \ottdefnATPInfLabeled{}
  \hfill \\
  \ottdefnATNInfLabeled{}
  \hfill \\
  \ottdefnATSpinInfLabeled{}
  \caption{Algorithmic Type Inference}
  \label{fig:type-inference}
\end{figure}

@@

\paragraph*{Application to an Empty List of Arguments}
  Rule \ruleref{\ottdruleATEmptyAppLabel}
  is the base case of application inference. 
  If the list of arguments is empty, 
  the inferred type is the type of the head,
  and the algorithm returns it after normalizing.

@@

To find this substitution, we use the minimal instantiation
algorithm (\cref{fig:minimal-instantiation}). In the judgment "$[[Γ ⊢ uP SC minby uσ]]$",
$[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
instantiation of $[[uP]]$ that satisfies $[[SC]]$. 
First, it removes the existential quantifiers by \ruleref{\ottdruleSINGExistsLabel}.
Then it considers two cases. 
\begin{enumerate}
  \item If the type $[[uP]]$ is an algorithmic variable $[[α̂⁺]]$
    restricted by a \emph{subtyping} constraint $[[(α̂⁺ :≥ iQ) ∊ SC]]$,
    its minimal instantiation is the type $[[nf(iQ)]]$, and 
    \ruleref{\ottdruleSINGPUvarLabel} is applied.
  \item Otherwise, the type $[[uP]]$ is either a shifted computation, a
    declarative variable, an unrestricted algorithmic variable, or an
    algorithmic variable restricted by an \emph{equivalence} constraint. In all
    of these cases, the minimal instantiation exists if and only if there is
    only one possible instantiation of $[[uP]]$ that satisfies $[[SC]]$.
